---
title: "Classification: Alternative Techniques"
---

## Install packages

Install the packages used in this chapter:

```{r setup_04, message=FALSE}
if(!require(pacman))
  install.packages("pacman")

pacman::p_load(
  C50,                # C5.0 Decision Trees and Rule-Based Models
  caret,              # Classification and Regression Training
  e1071,              # Misc Functions of the Department of Statistics (e1071), TU Wien
  keras,              # R Interface to 'Keras'
  kernlab,            # Kernel-Based Machine Learning Lab
  lattice,            # Trellis Graphics for R
  MASS,               # Support Functions and Datasets for Venables and Ripley's MASS
  mlbench,            # Machine Learning Benchmark Problems
  nnet,               # Feedforward Neural Networks and Multinomial Log-Linear Models
  palmerpenguins,     # Palmer Archipelago (Antarctica) Penguin Data
  party,              # A Laboratory for Recursive Partytioning
  partykit,           # A Toolkit for Recursive Partytioning
  randomForest,       # Breiman and Cutler's Random Forests for Classification and Regression
  rpart,              # Recursive partitioning models
  RWeka,              # R/Weka Interface
  scales,             # Scale Functions for Visualization
  tidymodels,         # Tidy machine learning framework
  tidyverse,          # Tidy data wrangling and visualization
  xgboost             # Extreme Gradient Boosting
)
```

Show fewer digits

```{r }
options(digits=3)
```

## Introduction

Many different [classification algorithms](https://en.wikipedia.org/wiki/Supervised_learning) have been proposed in the literature. In this chapter, we will apply some of the more popular methods.

## Training and Test Data

We will use the Zoo dataset which is included in the R package `mlbench` (you may have to install it). The Zoo dataset containing 17 (mostly logical) variables on different 101 animals as a data frame with 17 columns (hair, feathers, eggs, milk, airborne, aquatic, predator, toothed, backbone, breathes, venomous, fins, legs, tail, domestic, catsize, type). We convert the data frame into a tidyverse tibble (optional).

```{r }
data(Zoo, package="mlbench")
Zoo <- as.data.frame(Zoo)
Zoo |> glimpse()
```

We will use the package [**caret**](https://topepo.github.io/caret/) to make preparing training sets and building classification (and regression) models easier. A great cheat sheet can be found [here](https://ugoproto.github.io/ugo_r_doc/pdf/caret.pdf).

Multi-core support can be used for cross-validation. **Note:** It is commented out here because it does not work with rJava used in RWeka below.

```{r }
##library(doMC, quietly = TRUE)
##registerDoMC(cores = 4)
##getDoParWorkers()
```

Test data is not used in the model building process and needs to be set aside purely for testing the model after it is completely built. Here I use 80% for training.

```{r}
set.seed(123)  # for reproducibility
inTrain <- createDataPartition(y = Zoo$type, p = .8)[[1]]
Zoo_train <- dplyr::slice(Zoo, inTrain)
Zoo_test <- dplyr::slice(Zoo, -inTrain)
```

## Fitting Different Classification Models to the Training Data

Create a fixed sampling scheme (10-folds) so we can compare the fitted models later.

```{r }
train_index <- createFolds(Zoo_train$type, k = 10)
```

The fixed folds are used in `train()` with the argument `trControl = trainControl(method = "cv", indexOut = train_index))`. If you don't need fixed folds, then remove `indexOut = train_index` in the code below.

For help with building models in caret see: `? train`

**Note:** Be careful if you have many `NA` values in your data. `train()` and cross-validation many fail in some cases. If that is the case then you can remove features (columns) which have many `NA`s, omit `NA`s using `na.omit()` or use imputation to replace them with reasonable values (e.g., by the feature mean or via kNN). Highly imbalanced datasets are also problematic since there is a chance that a fold does not contain examples of each class leading to a hard to understand error message.

### Conditional Inference Tree (Decision Tree)

```{r }
ctreeFit <- Zoo_train |> train(type ~ .,
  method = "ctree",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
ctreeFit
plot(ctreeFit$finalModel)
```

### C 4.5 Decision Tree

```{r }
C45Fit <- Zoo_train |> train(type ~ .,
  method = "J48",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
C45Fit
C45Fit$finalModel
```

### K-Nearest Neighbors

**Note:** kNN uses Euclidean distance, so data should be standardized (scaled) first. Here legs are measured between 0 and 6 while all other variables are between 0 and 1. Scaling can be directly performed as preprocessing in `train` using the parameter `preProcess = "scale"`.

```{r }
knnFit <- Zoo_train |> train(type ~ .,
  method = "knn",
  data = _,
  preProcess = "scale",
	tuneLength = 5,
  tuneGrid=data.frame(k = 1:10),
	trControl = trainControl(method = "cv", indexOut = train_index))
knnFit

knnFit$finalModel
```

### PART (Rule-based classifier)

```{r }
rulesFit <- Zoo_train |> train(type ~ .,
  method = "PART",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv", indexOut = train_index))
rulesFit

rulesFit$finalModel
```

### Linear Support Vector Machines

```{r }
svmFit <- Zoo_train |> train(type ~.,
  method = "svmLinear",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
svmFit

svmFit$finalModel
```

### Random Forest

```{r }
randomForestFit <- Zoo_train |> train(type ~ .,
  method = "rf",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
randomForestFit

randomForestFit$finalModel
```

### Gradient Boosted Decision Trees (xgboost)

```{r }
xgboostFit <- Zoo_train |> train(type ~ .,
  method = "xgbTree",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv", indexOut = train_index),
  tuneGrid = expand.grid(
    nrounds = 20,
    max_depth = 3,
    colsample_bytree = .6,
    eta = 0.1,
    gamma=0,
    min_child_weight = 1,
    subsample = .5
  ))
xgboostFit

xgboostFit$finalModel
```

### Artificial Neural Network

```{r }
nnetFit <- Zoo_train |> train(type ~ .,
  method = "nnet",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index),
  trace = FALSE)
nnetFit

nnetFit$finalModel
```

## Comparing Models

Collect the performance metrics from the models trained on the same data.

```{r }
resamps <- resamples(list(
  ctree = ctreeFit,
  C45 = C45Fit,
  SVM = svmFit,
  KNN = knnFit,
  rules = rulesFit,
  randomForest = randomForestFit,
  xgboost = xgboostFit,
  NeuralNet = nnetFit
    ))
resamps
```

Calculate summary statistics

```{r }
summary(resamps)

library(lattice)
bwplot(resamps, layout = c(3, 1))
```

Perform inference about differences between models. For each metric, all pair-wise differences are computed and tested to assess if the difference is equal to zero. By default Bonferroni correction for multiple comparison is used. Differences are shown in the upper triangle and p-values are in the lower triangle.

```{r }
difs <- diff(resamps)
difs
summary(difs)
```

All perform similarly well except ctree (differences in the first row are negative and the p-values in the first column are \<.05 indicating that the null-hypothesis of a difference of 0 can be rejected).

## Applying the Chosen Model to the Test Data

Most models do similarly well on the data. We choose here the random forest model.

```{r }
pr <- predict(randomForestFit, Zoo_test)
pr
```

Calculate the confusion matrix for the held-out test data.

```{r }
confusionMatrix(pr, reference = Zoo_test$type)
```

## Comparing Decision Boundaries of Popular Classification Techniques

Classifiers create decision boundaries to discriminate between classes. Different classifiers are able to create different shapes of decision boundaries (e.g., some are strictly linear) and thus some classifiers may perform better for certain datasets. This page visualizes the decision boundaries found by several popular classification methods.

The following plot adds the decision boundary (black lines) and classification confidence (color intensity) by evaluating the classifier at evenly spaced grid points. Note that low resolution (to make evaluation faster) will make the decision boundary look like it has small steps even if it is a (straight) line.

```{r message=FALSE, warning=FALSE}
library(scales)
library(tidyverse)
library(ggplot2)
library(caret)

decisionplot <- function(model, data, class_var, 
  predict_type = c("class", "prob"), resolution = 3 * 72) {
  # resolution is set to 72 dpi if the image is rendered  3 inches wide. 
  
  y <- data |> pull(class_var)
  x <- data |> dplyr::select(-all_of(class_var))
  
  # resubstitution accuracy
  prediction <- predict(model, x, type = predict_type[1])
  # LDA returns a list
  if(is.list(prediction)) prediction <- prediction$class
  prediction <- factor(prediction, levels = levels(y))
  
  cm <- confusionMatrix(data = prediction, 
                        reference = y)
  acc <- cm$overall["Accuracy"]
  
  # evaluate model on a grid
  r <- sapply(x[, 1:2], range, na.rm = TRUE)
  xs <- seq(r[1,1], r[2,1], length.out = resolution)
  ys <- seq(r[1,2], r[2,2], length.out = resolution)
  g <- cbind(rep(xs, each = resolution), rep(ys, time = resolution))
  colnames(g) <- colnames(r)
  g <- as_tibble(g)
  
  ### guess how to get class labels from predict
  ### (unfortunately not very consistent between models)
  cl <- predict(model, g, type = predict_type[1])
  
  # LDA returns a list
  prob <- NULL
  if(is.list(cl)) { 
    prob <- cl$posterior
    cl <- cl$class
  } else
    if(!is.na(predict_type[2]))
      try(prob <- predict(model, g, type = predict_type[2]))
  
  # we visualize the difference in probability/score between the 
  # winning class and the second best class.
  # don't use probability if predict for the classifier does not support it.
  max_prob <- 1
  if(!is.null(prob))
    try({
      max_prob <- t(apply(prob, MARGIN = 1, sort, decreasing = TRUE))
      max_prob <- max_prob[,1] - max_prob[,2]
    }, silent = TRUE) 
  
  cl <- factor(cl, levels = levels(y))
  
  g <- g |> add_column(prediction = cl, probability = max_prob)
  
  ggplot(g, mapping = aes(
    x = .data[[colnames(g)[1]]], y = .data[[colnames(g)[2]]])) +
    geom_raster(mapping = aes(fill = prediction, alpha = probability)) +
    geom_contour(mapping = aes(z = as.numeric(prediction)), 
      bins = length(levels(cl)), linewidth = .5, color = "black") +
    geom_point(data = data, mapping =  aes(
      x = .data[[colnames(data)[1]]], 
      y = .data[[colnames(data)[2]]],
      shape = .data[[class_var]]), alpha = .7) + 
    scale_alpha_continuous(range = c(0,1), limits = c(0,1), guide = "none") +  
    labs(subtitle = paste("Training accuracy:", round(acc, 2))) +
     theme_minimal(base_size = 14)
}
```

### Penguins Dataset

For easier visualization, we use two dimensions of the `penguins` dataset. Contour lines visualize the density like mountains on a map.

```{r }
set.seed(1000)
data("penguins")
penguins <- as_tibble(penguins) |>
  drop_na()

### Three classes 
### (note: MASS also has a select function which hides dplyr's select)
x <- penguins |> dplyr::select(bill_length_mm, bill_depth_mm, species)
x

ggplot(x, aes(x = bill_length_mm, y = bill_depth_mm, fill = species)) +  
  stat_density_2d(geom = "polygon", aes(alpha = after_stat(level))) +
  geom_point() +
  theme_minimal(base_size = 14) +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)",
       fill = "Species",
       alpha = "Density")
```

*Note:* There is some overplotting and you could use `geom_jitter()` instead of `geom_point()`.

#### K-Nearest Neighbors Classifier

```{r }
model <- x |> caret::knn3(species ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "species") + 
  labs(title = "kNN (1 neighbor)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> caret::knn3(species ~ ., data = _, k = 3)
decisionplot(model, x, class_var = "species") + 
  labs(title = "kNN (3 neighbor)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> caret::knn3(species ~ ., data = _, k = 9)
decisionplot(model, x, class_var = "species") + 
  labs(title = "kNN (9 neighbor)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

Increasing $k$ smooths the decision boundary. At $k=1$, we see white areas around points where penguins of two classes are in the same spot. Here, the algorithm randomly chooses a class during prediction resulting in the meandering decision boundary. The predictions in that area are not stable and every time we ask for a class, we may get a different class.

#### Naive Bayes Classifier

```{r }
model <- x |> e1071::naiveBayes(species ~ ., data = _)
decisionplot(model, x, class_var = "species", 
             predict_type = c("class", "raw")) + 
  labs(title = "Naive Bayes",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction") 
```

#### Linear Discriminant Analysis

```{r }
model <- x |> MASS::lda(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "LDA",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

#### Multinomial Logistic Regression (implemented in nnet)

Multinomial logistic regression is an extension of logistic regression to problems with more than two classes.

```{r }
model <- x |> nnet::multinom(species ~., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "Multinomial Logistic Regression",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

#### Decision Trees

```{r }
model <- x |> rpart::rpart(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "CART",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> rpart::rpart(species ~ ., data = _,
  control = rpart.control(cp = 0.001, minsplit = 1))
decisionplot(model, x, class_var = "species") + 
  labs(title = "CART (overfitting)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> C50::C5.0(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "C5.0",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> randomForest::randomForest(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "Random Forest",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

#### SVM

```{r }
model <- x |> e1071::svm(species ~ ., data = _, kernel = "linear")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (linear kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> e1071::svm(species ~ ., data = _, kernel = "radial")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (radial kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> e1071::svm(species ~ ., data = _, kernel = "polynomial")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (polynomial kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <- x |> e1071::svm(species ~ ., data = _, kernel = "sigmoid")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (sigmoid kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

#### Single Layer Feed-forward Neural Networks

```{r }
model <-x |> nnet::nnet(species ~ ., data = _, size = 1, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (1 neuron)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <-x |> nnet::nnet(species ~ ., data = _, size = 2, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (2 neurons)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <-x |> nnet::nnet(species ~ ., data = _, size = 4, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (4 neurons)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")

model <-x |> nnet::nnet(species ~ ., data = _, size = 10, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (10 neurons)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

### Circle Dataset

This set is not linearly separable!

```{r }
set.seed(1000)

x <- mlbench::mlbench.circle(500)
###x <- mlbench::mlbench.cassini(500)
###x <- mlbench::mlbench.spirals(500, sd = .1)
###x <- mlbench::mlbench.smiley(500)
x <- cbind(as.data.frame(x$x), factor(x$classes))
colnames(x) <- c("x", "y", "class")
x <- as_tibble(x)
x

ggplot(x, aes(x = x, y = y, color = class)) + 
  geom_point() +
  theme_minimal(base_size = 14)
```

#### K-Nearest Neighbors Classifier

```{r }
model <- x |> caret::knn3(class ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (1 neighbor)",
       shape = "Class",
       fill = "Prediction")

model <- x |> caret::knn3(class ~ ., data = _, k = 10)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (10 neighbor)",
       shape = "Class",
       fill = "Prediction")
```

#### Naive Bayes Classifier

```{r }
model <- x |> e1071::naiveBayes(class ~ ., data = _)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + 
  labs(title = "naive Bayes",
       shape = "Class",
       fill = "Prediction")
```

#### Linear Discriminant Analysis

LDA cannot find a good model since the true decision boundary is not linear.

```{r }
model <- x |> MASS::lda(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "LDA",
       shape = "Class",
       fill = "Prediction")
```

#### Logistic Regression (implemented in nnet)

Multinomial logistic regression is an extension of logistic regression to problems with more than two classes. It also tries to find a linear decision boundary.

```{r }
model <- x |> nnet::multinom(class ~., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "Multinomial Logistic Regression",
       shape = "Class",
       fill = "Prediction")
```

#### Decision Trees

```{r }
model <- x |> rpart::rpart(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART",
       shape = "Class",
       fill = "Prediction")

model <- x |> rpart::rpart(class ~ ., data = _,
  control = rpart.control(cp = 0.001, minsplit = 1))
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART (overfitting)",
       shape = "Class",
       fill = "Prediction")

model <- x |> C50::C5.0(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "C5.0",
       shape = "Class",
       fill = "Prediction")

library(randomForest)
model <- x |> randomForest(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "Random Forest",
       shape = "Class",
       fill = "Prediction")
```

#### SVM

Linear SVM does not work for this data.

```{r }
model <- x |> e1071::svm(class ~ ., data = _, kernel = "linear")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (linear kernel)",
       shape = "Class",
       fill = "Prediction")

model <- x |> e1071::svm(class ~ ., data = _, kernel = "radial")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (radial kernel)",
       shape = "Class",
       fill = "Prediction")

model <- x |> e1071::svm(class ~ ., data = _, kernel = "polynomial")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (polynomial kernel)",
       shape = "Class",
       fill = "Prediction")

model <- x |> e1071::svm(class ~ ., data = _, kernel = "sigmoid")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (sigmoid kernel)",
       shape = "Class",
       fill = "Prediction")
```

#### Single Layer Feed-forward Neural Networks

```{r }
model <-x |> nnet::nnet(class ~ ., data = _, size = 1, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (1 neuron)",
       shape = "Class",
       fill = "Prediction")

model <-x |> nnet::nnet(class ~ ., data = _, size = 2, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (2 neurons)",
       shape = "Class",
       fill = "Prediction")

model <-x |> nnet::nnet(class ~ ., data = _, size = 4, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (4 neurons)",
       shape = "Class",
       fill = "Prediction")

model <-x |> nnet::nnet(class ~ ., data = _, size = 10, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (10 neurons)",
       shape = "Class",
       fill = "Prediction")
```

## More Information on Classification with R

-   Package caret: <http://topepo.github.io/caret/index.html>
-   Tidymodels (machine learning with tidyverse): <https://www.tidymodels.org/>
-   R taskview on machine learning: <http://cran.r-project.org/web/views/MachineLearning.html>
